---
title: "Unit6CaseStudy_TidyData"
author: "Oscar Padilla"
date: "June 13, 2016"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache= FALSE)
knitr::opts_chunk$set(echo = TRUE)
```

#CASE STUDY FOR UNIT 6
##Tidy Data
Let's start with the GDP file 

```{r}
library(repmis)
library(dplyr)
library(ggplot2)
library(reshape2)
library(countrycode)
library(RCurl)
GDPurl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv"
rawGDP <- source_data(GDPurl, sep = ",", header = FALSE, stringsAsFactors = TRUE)
#Eliminate header and footer non-variable information
rawGDP <- rawGDP[6:195, ]
#Select only valid variables
GDPSelected <- dplyr::select(rawGDP, V1, V2, V4, V5)
#Rename variables to be human readable
names(GDPSelected) <- c("country_short", "rank", "country_long", "GDP2012")
#Look for NA values
GDPSelected[is.na(GDPSelected$GDP2012), ]
#Convert GDP into a numeric variable
GDPSelected$GDP2012 <- as.numeric(gsub(",", "", GDPSelected$GDP2012))
#Sort the data frame in ascending order by GDP (so United States is last). What is the 13th country in the resulting data frame
GDPclean <- arrange(GDPSelected, GDP2012)
GDPclean <- rename(GDPclean, CountryCode = country_short)
str(GDPclean)
#Top ten economies in descending order (largest economy first)
arrange(tail(GDPclean, 10), desc(GDP2012))
#Last 13 countries by GDP size (smallest economy first)
head(GDPclean, 13)
```
###2. Sort the data frame in ascending order by GDP rank (so United States is last). What is the 13th country in the resulting data frame?
*The 13th country is `r GDPclean[13, 3]` with a GDP of $ `r GDPclean[13, 4]` million USD.*

Let's move on to the educational data set.

```{r}
EduURL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv"
rawEdu <- source_data(EduURL, sep = ",", header = TRUE, stringsAsFactors = TRUE)
#Select variables (columns) we actually need: first Country Code and third Income Group
EduSelected <- select(rawEdu, 1, 3)
names(EduSelected)[2] <- "IncomeGroup"
str(EduSelected)
```

Once we have cleaned up both files, let's proceed to merge both data sets.

```{r}
GDPEdu <- merge(GDPclean, EduSelected, by = "CountryCode", all = TRUE)
```

###1. Match the data based on the country shortcode. How many of the IDs match?
*However `r nrow(GDPEdu[is.na(GDPEdu$GDP2012), ])` countries are missing GDP data.* which will be eliminated.

###3. What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups?
```{r}
GDPEduSelected <- GDPEdu[!is.na(GDPEdu$GDP2012), ]
GDPEduclean <- arrange(GDPEduSelected, desc(GDP2012))
GDPEduclean$rank_num <- as.numeric(as.character(GDPEduclean$rank))
attach(GDPEduclean)
GDPEduclean[!is.na(GDPEduclean$IncomeGroup) & GDPEduclean$IncomeGroup == "High income: OECD", ]
# Average GDP rankings for the "High income: OECD" 
mean(GDPEduclean[!is.na(GDPEduclean$IncomeGroup) & GDPEduclean$IncomeGroup == "High income: OECD", 6])
```

```{r}
GDPEduclean[!is.na(GDPEduclean$IncomeGroup) & GDPEduclean$IncomeGroup == "High income: nonOECD", ]
# Average GDP rankings for the "High income: nonOECD" 
mean(GDPEduclean[!is.na(GDPEduclean$IncomeGroup) & GDPEduclean$IncomeGroup == "High income: nonOECD", 6])
```

###4.	Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.
```{r}
plot(GDPEduclean$rank_num, log10(GDP2012))
ggplot(data = GDPEduclean, aes(x = rank_num, y = log10(GDP2012))) + geom_point(aes(col = IncomeGroup)) + labs(title = "GDP as Function of Country Ranking", x = "GDP ranking", y = "log base 10 GDP")+ scale_fill_manual(breaks = 50, 100, 150, 200, 250)
```

